Applying gamification to developer practices

### 2013-09-05 paper_id1.tex
study design looking good

\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{
\alignauthor
Will Snipes
\affaddr{ABB Corporate Research}\\
\affaddr{Industrial Software Systems}\\
\affaddr{Raleigh, NC USA}\\
\email{will.snipes@us.abb.com}
\and
\alignauthor
Anil R. Nair
\affaddr{ABB Corporate Research}\\
\affaddr{Industrial Software Systems}\\
\affaddr{Bangalore, India}\\
\email{anil.nair@in.abb.com}
\and
\alignauthor
Emerson Murphy-Hill
\affaddr{North Carolina State University}\\
\affaddr{Department of Computer Science}\\
\affaddr{Raleigh, N.C. USA}\\
\email{emerson@csc.ncsu.edu}\\
}

@@

Utilizing the unobtrusive continuous monitoring capabilities of Blaze, we designed the pilot as a longtiudinal study in three waves.  Longtiudinal stuides allow us to design an experiement and determine the effects over two or more sample periods using standard MANOVA (Multiple ANalysis of VAriance) models. \cite{longitudinalbook}  We employed the three wave model defining the first wave as a baseline capturing the current practices of the pilot group.  The second wave employed the intervention that included an informative web site on structured navigation commands in Visual Studio, information about the sando search tool, and information about how points would be awarded during the pilot.  The third wave provided the participants with continuous feedback in the from of points accumulated and an indicator arrow \ref{fig:blazegui} that points up when their navigation ratio is higher than their historical average.

@@

\subsection{Setup the practice}

Selecting a practice to study involves defining specific characteristics of the practice and how capturing actions in Visual Studio relates to the practice.  In this study we selected the structured navigation practice shown by Robillard et. al. as correlated with effective program maintenance \cite{wbsnipes:Robillard2004How} .  Studying structured navigation is a natural fit for capturing commands issued in the IDE.  The first step is to create the classification of commands that are involved in navigation.  Then differentiate which commands we consider as structured navigation from unstructured navigation.

The Blaze tool provides an XML configuration file that allows defining multiple levels of categories for commands thus we can categorize commands that are navigation then further classify them into structured and unstructured navigation.
We classified commands as structured navigation when they allow the developer to locate a code element following the program structure from another element.  For example, Navigate To \textit{Ctrl+,}, Go To Definition \textit{F12}, and Find All References \textit{Ctrl+K,R} are commands that follow structure of the code that we classify as structured navigation.
Commands that we considered as unstructured navigation include scrolling through a file using keys or the scroll-bar, opening files from the Solution Explorer, Find commands (e.g. Find in Files), Go-to line number, and children of these commands (e.g. Find Next).

Another factor to consider is the design of achievement awards and points.  The design of achievement awards should gradually encourage the participant to reach higher and higher levels of success in the activity.  Achievements must feel earned to the participant so that internally they recognize the effort to get the award was significant and feel satisfied  \cite{wbsnipes:Hamari2011Framework}.  Hamari and Eranti describe a general relationship between achievements in games and regular game play  \cite{wbsnipes:Hamari2011Framework} .  How those achievements are awarded to players, and the concept of levels of achievements.  The article contains a warning that assigning achievements to tasks reduces intrinsic motivation because autonomy is reduced by the achievements.   In games achievements are typically a parallel scoring system to the main game play.  They make the game more engaging providing multiple ways to increase your score and multiple challenges in one interface.

Using these guidelines we assigned points to structured navigation commands which received 1-10 points each while unstructured navigation commands received 0 points.  Assigning negative points is discouraged as this may demotivate folks who do not have detailed knowledge of how the points are assigned.\cite{}   In creating levels, following the guidance from Hamari \cite{wbsnipes:Hamari2011Framework}, we built levels following an exponential curve.  Users initially level up after a working day or two extending to achieving the next level after a week or even a month following an exponential curve.  The leveling up points make the game motivating for beginners and challenging for more experienced developers.  We designed the points and levels such that above average developers can complete all levels during the study period.

\subsection{Study Intervention and Participants}

@@

After a week's data collection, Blaze started to automatically pop-up a window containing a button for a web page with information.  The page included a link to download another tool for code search called Sando \ref{sando} and information about using that tool in a video.  The sub-page on navigation contained a page of tutorial on the built-in structured navigation commands in Visual Studio.  The About Blaze page contained details of the point scoring system and background on Blaze's purpose.  The usefulness of the communication site is rated as part of our study completion survey.

We conducted this study with an intact team of developers working in an R\&D facility in India on a large industrial software system.  Developers were asked to volunteer for the pilot by their management, however, managers did not know who participated in the study.  We intended to keep managers out of the loop to reduce concerns among the developers that the data may be used for purposes other then the intended study.  By selecting an intact team, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition among teammates.

>>>>

\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{
\alignauthor
Will Snipes\\
\affaddr{ABB Corporate Research}\\
\affaddr{Industrial Software Systems}\\
\affaddr{Raleigh, NC USA}\\
\email{will.snipes@us.abb.com}
\and
\alignauthor
Anil R. Nair\\
\affaddr{ABB Corporate Research}\\
\affaddr{Industrial Software Systems}\\
\affaddr{Bangalore, India}\\
\email{anil.nair@in.abb.com}
\and
\alignauthor
Emerson Murphy-Hill
\affaddr{North Carolina State University}\\
\affaddr{Department of Computer Science}\\
\affaddr{Raleigh, N.C. USA}\\
\email{emerson@csc.ncsu.edu}\\
}

@@

Utilizing the unobtrusive continuous monitoring capabilities of Blaze, we designed the pilot as a longtiudinal study in three waves.  Longtiudinal studies allow us to design an experiment and determine the effects over two or more sample periods using standard MANOVA (Multiple ANalysis of VAriance) models. \cite{longitudinalbook}  We employed the three wave model defining the first wave as a baseline capturing the current practices of the pilot group.  The second wave employed the intervention that included an informative web site on structured navigation commands in Visual Studio, information about the Sando search tool, and information about how points would be awarded during the pilot.  The third wave provided the participants with continuous feedback in the from of points accumulated and an indicator arrow \ref{fig:blazegui} that points up when their navigation ratio is higher than their historical average.

@@

\subsection{Select a Practice}
In their paper on game design patterns, Hamari et. al. issue a warning that assigning achievements to required tasks reduces intrinsic motivation because player autonomy is reduced by the achievements \cite{wbsnipes:Hamari2011Framework} .  This warning indicates that we should avoid giving points for things that are already an expected deliverable such as finding or fixing bugs, completing development tasks, or meeting required process criteria.  Practices to consider are those practices with room to go beyond a checklist or minimum criteria where developers have the opportunity to excel.

Selecting a practice to study involves identifying a practice, defining specific characteristics of the practice to measure, and identifying the actions in Visual Studio that relate to the practice.  We currently limit ourselves to practices that exist entirely in the IDE thus can be monitored using Blaze.  We identified potential practices of test-driven development, debugging practices, navigation practices,  eliminating static analysis bugs, and check-in frequency.  Practices of check-in frequency, static analysis bug elimination are highlighted in this paper's related work section with prior gamification studies in classroom environments.  Johnson and Kou achieved automated monitoring of test-driven development practices with Zorro thus may provide a candidate for future gamification application.   Debugging practices have developer community opinions on optimal methods, however, detailed debug practices with modern IDEs have not received research community attention, thus are also reserved for future work.  Based on studies of structured navigation by Robillard et. al. that identify structured navigation as a practice leading to effective program maintenance  \cite{wbsnipes:Robillard2004How}, we identified structured navigation as a key practice.   Structured navigation fits the requirement of a practice that is contained in the IDE, is something that has proven benefits, and lends itself to measurement and scoring, is not an assigned task, and leaves room for developers to excel.

\subsection{Design the Game}

To setup the game for a practice, the Blaze tool provides an XML configuration file where the researcher can configure Blaze to categorize and assign points to the commands that are part of a software enginering practice.  Blaze allows the researcher to define multiple command category levels.  Thus we can categorize commands that are navigation then further classify them into structured and unstructured navigation.  We categorize commands as structured navigation when they allow the developer to locate a code element following the program structure from another element.  For example, Navigate To \textit{Ctrl+,}, Go To Definition \textit{F12}, and Find All References \textit{Ctrl+K,R} are commands that follow structure of the code, which we classify as structured navigation.
Commands that we considered as unstructured navigation include scrolling through a file using keys or the scroll-bar, opening files from the Solution Explorer, Find commands (e.g. Find in Files), Go-to line number, and children of these commands (e.g. Find Next).

Gamification practices consider the feedback of achievements and points as critical to a successful outcome.  The design of achievement awards should gradually encourage the participant to reach higher and higher levels of success in the activity.  Achievements must feel earned to the participant so that internally they recognize the effort to get the award was significant and feel satisfied  \cite{wbsnipes:Hamari2011Framework}.  Hamari and Eranti describe a general relationship between achievements in games and regular game play      In games achievements are typically a parallel scoring system to the main game play.  They make the game more engaging providing multiple ways to increase your score and multiple challenges in one interface.

Using these guidelines we assigned points to structured navigation commands which received 1-10 points each while unstructured navigation commands received 0 points.  Assigning negative points is discouraged as this may demotivate folks who do not have detailed knowledge of how the points are assigned.\cite{}   In creating levels, following the guidance from Hamari \cite{wbsnipes:Hamari2011Framework}, we built levels following an exponential curve.  Users initially level up after a working day or two extending to achieving the next level after a week or even a month following an exponential curve.  The levels make the game motivating for beginners and challenging for more experienced developers.  We designed the points and levels such that above average developers can complete all levels during the study period.

\subsection{Intervention Staging}

@@

After a week's data collection marking the end of the first wave, Blaze started to automatically pop-up a window containing a button for a web page with information.  The page included a link to download another tool for code search called Sando \ref{sando} and information about using that tool in a video.  The sub-page on navigation contained a page of tutorial on the built-in structured navigation commands in Visual Studio.  The About Blaze page contained details of the point scoring system and background on Blaze's purpose.  The usefulness of the communication site is rated as part of our study completion survey.

In the third wave developers received instant feedback from Blaze on the points they accumulated during the development session.  The up/down arrow provided a simple indicator on whether they were increasing their use of structured navigation or were using it less than their personal historical average.  This final configuration continued until the end of the pilot.

\subsection{Participants}

We conducted this study with an intact team of developers working in an R\&D facility in India on a large industrial software system.  Developers were asked to volunteer for the pilot by their management, however, managers did not know who participated in the study.  Mechanisms in our data collection to identify individuals were under the control of each developer, they could chose to share or keep confidential the association of a generated unique id with researchers or each other.

Management wished to have no knowledge of who was participating and took great care to remain detached from the pilot proceedings.  Data at this level of detail has not been collected in ABB development organizations except for a few minutes of video study that might be conducted.  Researchers were careful to create the system so that developers controlled their identification with the data and fortunately, management bought into the necessary anonymity of the participants.

\subsection{Competitive Elements}
By selecting an intact team, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition among teammates.    The leaderboard was presented so that developers could see their own score in relation to the top 5 people on the leaderboard.  However, the identifier of the other participants was generated as 3 capital letters assigned based on their position on the leaderboard.  Therefore the assigned letters could even change when an individual's ranking changed.  Not knowing who of your colleagues was ahead of you could be a factor in motivation, but also avoids one potential source of conflict in an intact team.

Competitive dynamics may be more effective when teams from different areas can compete as a team against other teams..  This tried and true method both builds esprit de corps and enhances achievement towards the intended outcome.  This study design is potential future work.
